---
title: "prepare data for catchment_scores"
author: "--"
date: "2/4/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#libraries
```{r}
library(sf)
library(tidygeocoder)
library(dplyr)
library(community)
```
#clean df
```{r}
#start clean
rm(list=ls())
```
#working directory
```{r}
setwd("~/VDH/Floating_Catchment_Areas/dmv/5. hospitals_dmv_model/")
```
#obtain data. Do not run 
```{r}
#a) conn

#b) query
provider <- sf::st_read(
  con, 
  query= "
SELECT  id , name, address, county, countyfips, type, latitude, longitude, beds, geoid_blk, state
FROM dc_health_behavior_diet.us_pl_hifld_2020_hospitals_geoblk A
WHERE (state = 'VA' OR state = 'MD' OR state = 'DC') 
AND  status='OPEN' 
AND (type != 'PSYCHIATRIC' OR type != 'MILITARY') "
)

#identify counties in NCR: National Capital Region
counties_ncr <- c(51013 , 51059  ,51600  ,51107  ,51610  ,51683  ,51685  ,51153  ,51510  ,11001  , 24021  ,24031  ,24033  ,24017) 

provider <- provider %>% filter(as.numeric(countyfips) %in% c(counties_ncr))

#c) Disconnect
RPostgreSQL::dbDisconnect(con)

#write
write.csv(provider, "hospitals_ncr.csv" )
```
# load data provider
```{r}
provider <- read.csv("hospitals_ncr.csv", row.names = 1)
provider <- provider %>% filter(type!= 'MILITARY')
provider <- provider %>% filter(name!="NAVAL HEALTH CLINIC (JOHN HENRY BALCH)")

names(provider)[7] <- 'lat'
names(provider)[8] <- 'lon'

## assign IDs just to be explicit
provider$ID <- paste0("l", seq_len(nrow(provider)))
```

# data combined
```{r}
## data combined
county_names <- list(
  dc = "District of Columbia",
  md = c("Charles", "Frederick", "Montgomery", "Prince George's"),
  va = c("Alexandria", "Arlington", "Fairfax", "Falls Church", "Loudoun", "Manassas", "Manassas Park", "Prince William")
)
data <- list()
shapes <- list()


for(state in c("dc", "md", "va")){
  # get county shapes to identify block groups
  counties <- download_census_shapes(".", state, "county", paste0(state, "_counties"))
  blockgroups <- download_census_shapes(".", state, "bg", paste0(state, "_blockgroups"))
  
  # save only selected counties
  counties <- counties[counties$NAME %in% county_names[[state]],]
  counties[counties$NAME == "Fairfax", "NAME"] <- c("Fairfax City", "Fairfax")
  shapes[[state]] <- list(
    counties = counties,
    blockgroups = blockgroups[substr(blockgroups$GEOID, 1, 5) %in% counties$GEOID,]
  )
  
  # population data
  data[[state]] <- download_census_population(".", state, 2019, counties = counties$GEOID)
}

data_combined <- do.call(rbind, lapply(names(data), function(state){
  d <- data[[state]]$estimates
  s <- shapes[[state]]$blockgroups
  rownames(s) <- s$GEOID
  total <- d$TOTAL.POPULATION_Total
  total[total == 0] <- 1
  data.frame(
    GEOID = d$GEOID,
    population = d$TOTAL.POPULATION_Total,
    percent_female = d$SEX.BY.AGE_Female_Female / total * 100,
    percent_white = d$RACE_Total_White.alone / total * 100,
    percent_over_49 = rowSums(d[, grep("[5-8][05]", colnames(d))]) / total * 100,
    female_over_14 = rowSums(d[, grep("Female_(?:[2-9]\\d|1[58])", colnames(d))]),
    pediatric_pop = rowSums(d[, c(4:7,28:31) ]),
    
    st_coordinates(st_centroid(st_geometry(s[as.character(d$GEOID),])))
  )
}))
data_combined <- data_combined[!is.na(data_combined$Y),]
```
## travel time
```{r}
library(osrm)
options(osrm.server = "http://104.248.112.16:5000/", osrm.profile = "car")
if(!file.exists("traveltimes_exercise.csv")){
  traveltimes <- osrmTable(
    src = data_combined[, c("GEOID", "X", "Y")],  #population-demand
    dst = provider[, c("ID", "lon", "lat")]     #providers supply
  )$duration
  write.csv(
    cbind(GEOID = rownames(traveltimes), as.data.frame(traveltimes)),
    "traveltimes_exercise.csv", row.names = FALSE
  )
}
```

# add1. Define geography id. This is because the Geography-GEOID from initial file may be outdated
```{r}
library(tigris)
library(maps)
# add block geoids
# get US blocks shapefile
blocks_VA <- st_as_sf(blocks(state="VA"))
blocks_MD <- st_as_sf(blocks(state="MD"))
blocks_DC <- st_as_sf(blocks(state="DC"))
blocks <- rbind(blocks_VA, blocks_MD)
blocks <- rbind(blocks, blocks_DC)

# lon and lat to geo-points
geopts <- provider %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4269)
# indeces of bgs which contain a geopoint
inds <- st_within(geopts$geometry, blocks$geometry, sparse=T)
blk_list <- c()
for (i in inds){
  if (identical(blocks$NAME[i],character(0))){
    blk_list<- append(blk_list, NA)}
  else{
    blk_list <- append(blk_list, blocks$GEOID[i])}
}
provider['geoid_blk_new'] <- blk_list


# This assumes that you have longitude and latitude of each location and if you want bg or tract or county just replace 'blocks_VA <- st_as_sf(blocks(state="VA"))' with a corresponding value, e.g. 'tracts_VA <- st_as_sf(tracts(state="VA"))'. I believe it uses either tigris or maps package

```

# add2 count providers per geography using matching codes
```{r}
#obtain geoid

provider$GEOID <- substr(provider$geoid_blk_new, 1,12)
data_combined$GEOID <- as.character(data_combined$GEOID)

#number of providers expanded
num_providers_wide <- full_join(provider %>% select(GEOID, beds ), data_combined %>% select(GEOID), by = "GEOID" )

#number providers by GEOID
hosp_cnt <- num_providers_wide %>% group_by(GEOID) %>% na.omit(name) %>% summarise(hosp_cnt = n( ), beds_cnt=sum(beds) )

#join hosp_cnt to geographies
data_combined <- data_combined %>% left_join(hosp_cnt, by= "GEOID")
```
# add3 mean and median of 10 nearest drive times 
```{r}
#mean of 10 nearest
top_mean <- function(x) {  
   mean(head(sort(x ), 10) ) }
#median of 10 nearest
top_median <- function(x) {  
   median(head(sort(x ), 10) ) }
#apply rowwise
traveltimes_near <- data.frame(near_10_mean=apply(traveltimes, 1, top_mean), 
                               near_10_median=apply(traveltimes, 1, top_median)) 
#rownames_to_column(traveltimes_near, var = "GEOID")
traveltimes_near$GEOID <- row.names(traveltimes_near) 
#join mean median traveltimes to geographies
data_combined <- data_combined %>% left_join(traveltimes_near, by= "GEOID")
```
#prepara data for save: SELECT relevant variables. 
```{r}
#raw traveltimes: traveltimes matrix already estimated and with colnames arranged 
traveltimes <- read.csv("traveltimes_exercise.csv", row.names = 1)
#population: always recheck relevant population: ie. for pediatrics: pop 0-17 years
population <- data_combined %>% select(GEOID, population, hosp_cnt, beds_cnt, near_10_mean, near_10_median)
# realign travel times
traveltimes <- traveltimes[as.character(population$GEOID), provider$ID]

```
# save new data
```{r}
write.csv(provider[, c("ID", "address", "lat", "lon", "beds")], "provider.csv", row.names = FALSE)
write.csv(cbind(GEOID = rownames(traveltimes), traveltimes), "traveltimes_trimmed.csv", row.names = FALSE)
write.csv(population, "population.csv", row.names = FALSE)
```



